---
title: "DNA/RNA Project – Final Report"
author: Group 4
date: May/June 2025
output:
  html_document

---

[Supplementary materials](https://github.com/Martinaa1408/DNARNA_Group4/blob/main/DNAmethylation_analysis_manual.pdf)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pipeline steps

**1. Load raw data with minfi and create an object called RGset storing the RGChannelSet object**  
  
`minfi` is an R package designed for the in-depth analysis of DNA methylation data obtained through Infinium microarrays. It provides tools for importing raw data, performing quality control, normalization, and a wide range of downstream analyses.  
Within the package, the RGChannelSet object represents a data structure that stores the signal intensities from the red and green channels of Illumina microarrays, forming a fundamental basis for quantifying methylation levels across the genome.  
`BiocManager` is the official package manager for Bioconductor packages. It automatically selects the Bioconductor repository compatible with our version of R and installs packages from Bioconductor (such as `minfi`), ensuring consistency across releases.  
  
First, check whether the `BiocManager` package is installed; if not, install it.  

```{r, message=FALSE, warning=FALSE, results='hide'}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
```

Install the `minfi` package using `BiocManager`, if it's not already available.

```{r, message=FALSE, warning=FALSE, results='hide'}
BiocManager::install("minfi")

library(minfi)
library(knitr)
```

Clean the R environment by removing all existing objects and set the working directory, where the raw data are stored.

```{r}
rm(list = ls())
directory = "C:/Users/coric/Desktop/1 YEAR/2 SEMESTER/DNA-RNA/PROGETTO/Final_Report-20250526/"
setwd(directory)
```

Load sample sheet using `read.metharray.sheet()`

```{r}
baseDir <- (directory)
targets <- read.metharray.sheet(baseDir)
# verify and print the number of imported samples
paste("Number of found samples: ", nrow(targets))
```

Read the data using the function `read.metharray.exp` and save it.

```{r}
RGset <- read.metharray.exp(targets = targets)
save(RGset, file = "RGset.RData")
```

**2. Create the dataframes Red and Green to store the red and green fluorescences respectively**  
  
In this phase of the analysis, extract the red (Cy5) and green (Cy3) fluorescence signals from the RGSet object, which contains the raw data from an Illumina microarray experiment. This is achieved using the `getRed()` and `getGreen()` functions from the `minfi` package, which provide access to the red and green signal channels, respectively.  
The extracted signals are converted into two separate data frames, named Red and Green, to facilitate data inspection, manipulation, and analysis.  
In each data frame, rows correspond to probes, and columns correspond to samples.  
  
To validate the extraction process, use the `dim()` and `head()` functions. `dim()` checks the dimensions of the matrices, and `head()` displays the first few rows. 

```{r, results='hide'}
Red <- data.frame(getRed(RGset))
dim(Red)
head(Red)

Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)
```

**3. Fill the following table: what are the Red and Green fluorescences for the address assigned to your group?**\
Optional: check in the manifest file if the address corresponds to a Type I or a Type II probe and, in case of Type I probe, report its color.\
  
In this step, we extract fluorescence intensities from the Red and Green channels for the probe identified by address "44666390".  

```{r}
address <- "44666390"
probe_red <- Red[address, ]
probe_green <- Green[address, ]
```

Load the Illumina manifest (*Illumina450Manifest_clean.RData*), in which control probes and SNP-associated probes (those prefixed with “rs”) have been removed to avoid ambiguity due to polymorphisms.  

```{r}
load(file.path(directory, "Illumina450Manifest_clean.RData"))
# Check if the address is in AddressA_ID or AddressB_ID
Illumina450Manifest_clean[Illumina450Manifest_clean$AddressA_ID=="44666390",'Infinium_Design_Type']
Illumina450Manifest_clean[Illumina450Manifest_clean$AddressB_ID=="44666390",'Infinium_Design_Type']
```

Create a summary dataframe for the selected address (B), combining the Red and Green fluorescence values for each sample and including the probe type.

```{r}
df_address <- data.frame(Sample=colnames(probe_green), Red_fluor=unlist(probe_red, use.names = FALSE ), Green_fluor=unlist(probe_green, use.names = FALSE), Type = "I")
knitr::kable(
  df_address,
  col.names = c("Sample", "Red fluor", "Green fluor", "Type"),
  align = "lrrc"
)
```

**4. Create the object MSet.raw**  
  
Convert the RGChannelSet to an MSet object using `preprocessRaw`.  
This function takes the Red and the Green channel of an Illumina methylation array, together with its associated manifest object and converts it into a MethylSet containing the methylated and unmethylated signal.

```{r, message=FALSE, warning=FALSE, results='hide'}
MSet.raw <- preprocessRaw(RGset)
MSet.raw
save(MSet.raw,file="MSet_raw.RData")
```

**5. Perform the following quality checks and provide a brief comment to each step:**

-   **QCplot**

-   **check the intensity of negative controls using minfi**

-   **calculate detection p-values for each sample, how many probes have a detection p-value higher than the threshold of 0.01?**

In a QCplot, samples methylation and unmethylation median signals are plotted. Good quality samples are identified in the upper-right section of the graph.  
  
**QCplot**

```{r, message=FALSE, warning=FALSE, results='hide'}
qc <- getQC(MSet.raw)
qc
```

```{r}
plotQC(qc)
```

Samples 1,3,8 are found slightly outside the threshold line, while samples 2 and 5 raise more concerns as they are far from optimal.  
  
Check the intensity of negative controls using `minfi`.

```{r, message=FALSE, warning=FALSE}
controlStripPlot(RGset, controls="NEGATIVE")
```

calculate detection p-values for each sample, how many probes have a detection p-value higher than the threshold of 0.01?

```{r}
threshold <- 0.01
detP <- detectionP(RGset)
failed <- detP> threshold

num_failed = colSums(failed)
df_failed <- data.frame(Sample=colnames(probe_green), Failed_Position=num_failed, perc_failed_probes=(means_of_columns <- colMeans(failed))*100, row.names = NULL)

knitr::kable(
  df_failed,
  col.names = c("Sample", "Failed Position", "% Failed Probes"),
  align = "lrc",
  digits = 2
)

```

Here are plotted all samples that failed the significance test, as their p-value was over the set  threshold (0.01%). This means that the signal captured by these probes cannot be distinguished from noise. From this dataframe is possible to highlight 4 different categories:  
-   High quality: sample 1 and 7 (less than 0.01% failed probes)  
-   Good quality: sample 2,4 and 6 (less than 0.2% failed probes)  
-   Low quality: sample 3 and 8 (above 0.2% failed probes)  
-   Critical: sample 5 (nearly 1% of failed probes)  
  
Given the results it might be appropriate to exclude sample 5 from the research, since it failed the QCplot and it possesses a considerable amount of failed probes.  
  
**6. Calculate raw beta and M values and plot the densities of mean methylation values, dividing the samples in CTRL and DIS (suggestion: subset the beta and M values matrixes in order to retain CTRL or DIS subjects and apply the function mean to the 2 subsets). Do you see any difference between the two groups?**  
  
**Beta-values** and **M-values** are complementary measures of DNA methylation: beta-values quantify the proportion of methylated signal over the total (0–1 range), while M-values express the log2 ratio of methylated to unmethylated intensities, offering better statistical properties for differential analysis.  
  
Select groups (CTRL and DIS) from targets.  

```{r}
ctrl <- basename(targets[targets$Group == "CTRL", "Basename"])
dis <- basename(targets[targets$Group == "DIS", "Basename"])
```

Remove any suffixes (e.g., .idat).

```{r}
ctrl <- gsub("\\.idat$", "", ctrl)
dis <- gsub("\\.idat$", "", dis)
```

Print the group samples for verification.

```{r}
cat("CTRL samples:", ctrl, "\n")
cat("DIS samples:", dis, "\n")
```

Suset the MSet object by groups.

```{r}
ctrlSet <- MSet.raw[, colnames(MSet.raw) %in% ctrl]
disSet <- MSet.raw[, colnames(MSet.raw) %in% dis]
```

Calculate Beta and M values.

```{r}
ctrlBeta <- getBeta(ctrlSet)
ctrlM <- getM(ctrlSet)
disBeta <- getBeta(disSet)
disM <- getM(disSet)
```

Calculate mean values per probe across samples.

```{r}
mean_ctrlBeta <- apply(ctrlBeta, 1, mean, na.rm = TRUE)
mean_disBeta <- apply(disBeta, 1, mean, na.rm = TRUE)
mean_ctrlM <- apply(ctrlM, 1, mean, na.rm = TRUE)
mean_disM <- apply(disM, 1, mean, na.rm = TRUE)
```


Density of mean Beta and M values.

```{r}
par(mfrow = c(1, 2))

# Density of mean Beta values
plot(density(mean_ctrlBeta, na.rm = TRUE),
     main = "Density of Beta Values",
     col = "#0067E6", lwd = 2.5,
     xlab = "Mean Beta Values", ylab = "Density")
lines(density(mean_disBeta, na.rm = TRUE), col = "#E50068", lwd = 2.5)
legend("topright", legend = c("CTRL", "DIS"), col = c("#0067E6", "#E50068"), lwd = 2)

# Density of mean M values
plot(density(mean_ctrlM, na.rm = TRUE),
     main = "Density of M Values",
     col = "#0067E6", lwd = 2.5,
     xlab = "Mean M Values", ylab = "Density")
lines(density(mean_disM, na.rm = TRUE), col = "#E50068", lwd = 2.5)
legend("topright", legend = c("CTRL", "DIS"), col = c("#0067E6", "#E50068"), lwd = 2)
```

Density of Beta values: The graph shows a bimodal distribution, that is to be expected for DNA methylation analysis. The first peak (β ~ 0.1) explains the proportion of unmethylated sites, while the peak on the right (β ~ 0.9) displays methylated sites.  
  
Density of M values: Bimodal symmetric distribution. The two peaks explain the proportion of unmethylated (M ~ -2) and methylated sites(M ~ +2).  
  
There are no clear differences between the control and disease group. This means that globally the methylation level distribution is the same, which is to be expected, as disease associated sites are localized in most cases.  
  
Reset plotting layout.

```{r}
par(mfrow = c(1, 1))
```

**7. Normalize the data using the function assigned to each group and compare raw data and normalized data. Produce a plot with 6 panels in which, for both raw and normalized data, you show the density plots of beta mean values according to the chemistry of the probes, the density plot of beta standard deviation values according to the chemistry of the probes and the boxplot of beta values. Provide a short comment about the changes you observe.**\
Optional: do you think that the normalization approach that you used is appropriate considering this specific dataset? Try to color the boxplots according to the group (CTRL and DIS) and check whether the distribution of methylation values is different between the two groups, before and after normalization.  
  
Subset the manifest by probe chemistry (Type I and Type II).

```{r}
dfI <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type == "I", ]
dfII <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type == "II", ]
dfI <- droplevels(dfI)
dfII <- droplevels(dfII)
```

Subset raw beta values by chemistry.

```{r}
beta_raw <- getBeta(MSet.raw)
beta_I_raw <- beta_raw[rownames(beta_raw) %in% dfI$IlmnID, ]
beta_II_raw <- beta_raw[rownames(beta_raw) %in% dfII$IlmnID, ]
```

Calculate mean and sd for raw Beta.

```{r}
mean_beta_I_raw <- rowMeans(beta_I_raw, na.rm = TRUE)
mean_beta_II_raw <- rowMeans(beta_II_raw, na.rm = TRUE)
sd_beta_I_raw <- apply(beta_I_raw, 1, sd, na.rm = TRUE)
sd_beta_II_raw <- apply(beta_II_raw, 1, sd, na.rm = TRUE)
```

Normalize using `preprocessFunnorm`.  
**Functional normalization** (FunNorm) is a between-array normalization method for Illumina 450K arrays that removes unwanted technical variation by regressing out the variability captured by control probes present on the array.

```{r, message=FALSE, warning=FALSE, results='hide'}
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
MSet.norm <- preprocessFunnorm(RGset)
beta_norm <- getBeta(MSet.norm)
```

Subset normalized Beta by chemistry.

```{r}
beta_I_norm <- beta_norm[rownames(beta_norm) %in% dfI$IlmnID, ]
beta_II_norm <- beta_norm[rownames(beta_norm) %in% dfII$IlmnID, ]
```

Calculate mean and sd for normalized Beta.

```{r}
mean_beta_I_norm <- rowMeans(beta_I_norm, na.rm = TRUE)
mean_beta_II_norm <- rowMeans(beta_II_norm, na.rm = TRUE)
sd_beta_I_norm <- apply(beta_I_norm, 1, sd, na.rm = TRUE)
sd_beta_II_norm <- apply(beta_II_norm, 1, sd, na.rm = TRUE)
```

Density plots for normalized Beta values.

```{r}
density_mean_beta_I_norm <- density(na.omit(mean_beta_I_norm))
density_mean_beta_II_norm <- density(na.omit(mean_beta_II_norm))
density_sd_beta_I_norm <- density(na.omit(sd_beta_I_norm))
density_sd_beta_II_norm <- density(na.omit(sd_beta_II_norm))
```

Prepare the colors for the groups.

```{r}
sample_prefixes <- sapply(strsplit(colnames(beta_raw), "_"), `[`, 1)
group_colors <- ifelse(sample_prefixes %in% targets[targets$Group == "CTRL", "SampleID"], "#0067E6", "#E50068")
```

Plotting: 2 rows x 3 columns.

```{r}
par(mfrow = c(2, 3))
# RAW DATA (first row)
# 1. Raw mean Beta densities
plot(density(mean_beta_I_raw, na.rm = TRUE), main = "Raw Beta Mean", col = "blue", lwd = 2, xlab = "Mean Beta")
lines(density(mean_beta_II_raw, na.rm = TRUE), col = "red", lwd = 2)
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lwd = 2)

# 2. Raw sd Beta densities
plot(density(sd_beta_I_raw, na.rm = TRUE), main = "Raw Beta SD", col = "blue", lwd = 2, xlab = "Beta SD")
lines(density(sd_beta_II_raw, na.rm = TRUE), col = "red", lwd = 2)
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lwd = 2)

# 3. Raw Beta boxplot (colored by group)
boxplot(beta_raw, outline = FALSE, col = group_colors,
        main = "Raw Beta Boxplot", ylab = "Beta Values", names = rep("", ncol(beta_raw)))
legend("topright", legend = c("CTRL", "DIS"), fill = c("#0067E6", "#E50068"))

# NORMALIZED DATA (second row)
# 4. Normalized mean Beta densities
plot(density(mean_beta_I_norm, na.rm = TRUE), main = "Norm. Beta Mean", col = "blue", lwd = 2, xlab = "Mean Beta")
lines(density(mean_beta_II_norm, na.rm = TRUE), col = "red", lwd = 2)
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lwd = 2)

# 5. Normalized sd Beta densities
plot(density(sd_beta_I_norm, na.rm = TRUE), main = "Norm. Beta SD", col = "blue", lwd = 2, xlab = "Beta SD")
lines(density(sd_beta_II_norm, na.rm = TRUE), col = "red", lwd = 2)
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lwd = 2)

# 6. Normalized Beta boxplot (colored by group)
boxplot(beta_norm, outline = FALSE, col = group_colors,
        main = "Norm. Beta Boxplot", ylab = "Beta Values", names = rep("", ncol(beta_norm)))
legend("topright", legend = c("CTRL", "DIS"), fill = c("#0067E6", "#E50068"))
```

The raw distributions of Beta mean and Beta standard deviation are clearly distinct between Type I and Type II probes. This can be explained by the chemistry implied in each probe type, which is different.  
The functional normalization clearly reduced the variability between type I and type II probes, using control intensities to systematically correct data.  
The box plot graph results are more uniform, while still preserving sample-specific variability.  
  
Reset plotting layout.

```{r}
par(mfrow = c(1, 1))
```

**8. Perform a PCA on the matrix of normalized beta values generated in step 7, after normalization.**  
**Comment the plot (Do you see any outlier?**  
**Do the samples divide according to the group?**  
**Do they divide according to the sex of the samples?**  
**Do they divide according to the batch, that is the column Sentrix_ID?).**  
  
**Principal Component Analysis** (PCA) is an unsupervised technique used to reduce data dimensionality while preserving key variability. It transforms correlated variables into uncorrelated principal components (PCs), allowing the identification of patterns and the main sources of variance.  
In this context, PCA is applied to evaluate whether CTRL and DIS samples show clear separation based on their gene expression profiles, potentially indicating biological differences.  
A scree plot displays the variance explained by each component and guides the selection of relevant PCs.  
  
Install required packages:  
`ggplot2` – implements the grammar of graphics for customizable PCA visualizations.  
`ggpubr` – provides simplified functions for generating publication-quality plots.
`cluster` – includes clustering algorithms and validation indices  
`factoMineR` – performs multivariate methods including PCA.  
`factoextra` – extracts and visualizes results from PCA and clustering.

```{r, warning=FALSE, message=FALSE, results='hide'}
pca_results <- prcomp(t(beta_norm), scale = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages(c("ggplot2", "ggpubr", "cluster", "factoMineR"))
install.packages("factoextra")
library(factoextra)
fviz_eig(pca_results, addlabels = TRUE, xlab = 'PC number', ylab = '% of variance', barfill = "#0063A6", barcolor = "black")
```

The scree plot shows that PC1 (29.8%) and PC2 (22.9%) together explain over 50% of the total variance. This means that the first two principal components capture the main structure in the normalized beta matrix, and are suitable for interpreting the most relevant trends in the data.  
  
**PCA Plot by Group**

```{r}
targets$Group <- as.factor(targets$Group)
palette(c("#E50068", "#0067E6"))
xlim <- range(pca_results$x[, 1])
ylim <- range(pca_results$x[, 2])
par(mar = c(5, 4, 4, 6), xpd = TRUE)
plot(pca_results$x[, 1], pca_results$x[, 2],
     cex = 1.2, pch = 19, col = targets$Group,
     xlab = "PC1", ylab = "PC2",
     xlim = xlim, ylim = ylim,
     main = "PCA (Groups)")
text(pca_results$x[, 1], pca_results$x[, 2],
     labels = rownames(pca_results$x), cex = 0.4, pos = 3)
legend("right", inset = c(-0.05, 0),
       legend = levels(targets$Group),
       col = c(1, 2), pch = 19,
       cex = 0.9, pt.cex = 1.2, x.intersp = 1.2,
       box.lty = 0, bg = "white")
```

CTRL and DIS samples tend to separate along PC1, with CTRL mainly on the left and DIS on the right. Although some overlap is present, the experimental group contributes to a visible variance pattern. No strong outliers are detected.  
  
**PCA plot by Sex**

```{r}
targets$Sex <- as.factor(targets$Sex)
palette(c("#C364CA", "#8BC4F9"))
par(mar = c(5, 4, 4, 6), xpd = TRUE)
plot(pca_results$x[, 1], pca_results$x[, 2],
     cex = 1.2, pch = 19, col = targets$Sex,
     xlab = "PC1", ylab = "PC2",
     xlim = xlim, ylim = ylim,
     main = "PCA (Sex)")
text(pca_results$x[, 1], pca_results$x[, 2],
     labels = rownames(pca_results$x), cex = 0.4, pos = 3)
legend("right", inset = c(-0.05, 0),
       legend = levels(targets$Sex),
       col = 1:nlevels(targets$Sex), pch = 19,
       cex = 0.9, pt.cex = 1.2, x.intersp = 1.2,
       box.lty = 0, bg = "white")
```

Male and female samples are mixed across PC1 and PC2. While minor variation exists along PC2, sex does not appear to be a major factor influencing the principal components. One male sample clusters close to a female cluster, showing no batch- or sex-driven structure.  
  
**PCA plot by Batch (Sentrix_ID)**

```{r}
targets$Slide <- as.factor(targets$Slide)
palette(c("#9e0059", "green"))
par(mar = c(5, 4, 4, 6), xpd = TRUE)
plot(pca_results$x[, 1], pca_results$x[, 2],
     cex = 1.2, pch = 19, col = targets$Slide,
     xlab = "PC1", ylab = "PC2",
     xlim = xlim, ylim = ylim,
     main = "PCA (Batch)")
text(pca_results$x[, 1], pca_results$x[, 2],
     labels = rownames(pca_results$x), cex = 0.4, pos = 3)
legend("right", inset = c(-0.05, 0),
       legend = levels(targets$Slide),
       col = c(1, 2), pch = 19,
       cex = 0.9, pt.cex = 1.2, x.intersp = 1.2,
       box.lty = 0, bg = "white")
```

Samples from different Sentrix_ID batches do not form distinct clusters. The PCA plot shows that normalization was effective in removing batch-associated variance. The distribution is scattered and intermixed, indicating a successful correction of technical variation.  
  
**9.	Using the matrix of normalized beta values generated in step 7, identify differentially methylated probes between group CTRL and group DIS using the function assigned to each group.**  
  
A statistical comparison was performed on the normalized beta values. t-test was applied to each probe (corresponding to a CpG site) by iterating over the rows of the normalized beta value matrix, using the `t.test()` function in R.  
The t-test allows comparison of group means while accounting for within-group variance, making it suitable for detecting site-specific differences.

```{r}
My_t_test <- function(x) {
  t_test <- t.test(x ~ targets$Group)
  return(t_test$p.value)
}

p_values <- apply(beta_norm, 1, My_t_test)
```

The `genefilter` package is used to perform statistical tests on each row of a data matrix, where rows typically represent features and columns represent samples. Specifically, the `rowttests` function efficiently applies Welch's t-test to each row, comparing two predefined groups.

```{r, warning=FALSE, message=FALSE, results='hide'}
if (!requireNamespace("genefilter", quietly = TRUE)) {
  BiocManager::install("genefilter")
}
library(genefilter)
# Create factor for the groups 
group_factor <- factor(targets$Group)
# Apply vectorized t-test  
t_test_results <- rowttests(beta_norm, group_factor)
# Extract only the p-values
p_values <- t_test_results$p.value
```

Create a dataframe combining beta values with the raw p-values.

```{r}
final_ttest <- data.frame(beta_norm, p_values_ttest = p_values)
```

Sort the probes by p-value in ascending order.

```{r}
final_ttest <- final_ttest[order(final_ttest$p_values_ttest), ]
```

Filter significant probes with p-value <= 0.05.

```{r}
final_ttest_th <- final_ttest[final_ttest$p_values_ttest <= 0.05, ]
```

Plot histogram of p-value distribution.

```{r}
hist(final_ttest$p_values_ttest, main = "P-value distribution (t-test)", 
     xlab = 'p-value')
abline(v = 0.05, col = "#ef233c")
```

There is a noticeable accumulation of low p-values, meaning that many probes have values close to 0. This pattern deviates from the uniform distribution expected under the null hypothesis and suggests that genuine differences in methylation exist between the two groups. The total number of significant probes was reported before applying multiple testing correction.    
  
Check the number of differentially methylated probes before correction.

```{r}
dim(final_ttest_th)
```

**10.	Apply multiple test correction and set a significant threshold of 0.05. How many probes do you identify as differentially methylated considering nominal pValues? How many after Bonferroni correction? How many after BH correction?**  
  
When multiple tests are performed simoultaneously, the likelihood of obtaining type 1 errors. This problem is mitigated using two different approaches:  
-   **Benjamini Hochberg** correction to control FDR (False Discovery Rate), which correspond to the expected proportion of type 1 errors among significant results;  
-   **Bonferroni correction** to control FWER (Family-Wise Error Rate), which correspond to the probability of obtaining type 1 errors among all the hypothesis.  
  
To apply these two types of correction, we will use `p.adjust()` function. The output is stored in the dataframe.
```{r}
corr_pval_BH <- p.adjust(final_ttest$p_values_ttest, method = "BH")  # Benjamini-Hochberg
corr_pval_bonferroni <- p.adjust(final_ttest$p_values_ttest, method = "bonferroni")
```

Combine the corrected p-values into the final dataframe for downstream filtering and visualization.

```{r}
final_ttest_corr <- data.frame(final_ttest, corr_pval_BH, corr_pval_bonferroni)
```

Count the number of differentially methylated probes before and after correction.

```{r}
before_correction <- nrow(final_ttest_corr[final_ttest_corr$p_values_ttest <= 0.05, ])
after_Bonferroni <- nrow(final_ttest_corr[final_ttest_corr$corr_pval_bonferroni <= 0.05, ])
after_BH <- nrow(final_ttest_corr[final_ttest_corr$corr_pval_BH <= 0.05, ])
```

Create a summary dataframe with all the results.

```{r}
diff_meth_df <- data.frame(before_correction, after_Bonferroni, after_BH)
rownames(diff_meth_df) <- c("differentially methylated probes")
```

Print the summary dataframe.

```{r}
knitr::kable(diff_meth_df)
```

Reset plotting area to a single plot (1 row x 1 column).

```{r}
par(mfrow = c(1,1))
```

Display boxplots of raw and adjusted p-values.

```{r}
boxplot(final_ttest_corr[,9:11], 
        col = c("#7038FF", "#C7FF38", "seashell2"),
        names = c("Raw", "BH", "Bonferroni"),
        main = 'P-value distribution (raw and adjusted)',
        ylab = "P-value")

legend("topright", 
       legend = c("raw", "BH", "Bonferroni"),
       col = c("#7038FF", "#C7FF38", "seashell2"),
       pch = 19, cex = 0.5, xpd = TRUE)
```

The box plot shows the distribution of p-values before and after correction. In this case, after applying both Bonferroni and BH corrections, no probes remained significant.  
  
**11.	Produce a volcano plot and a Manhattan plot of the results of differential methylation analysis**  

The **volcano plot** illustrates the outcome of the differential methylation analysis between diseased (DIS) and control (CTRL) groups. Each point in the plot corresponds to a CpG site tested for a significant difference in methylation (**DMPs**).  
The x-axis shows the difference in average beta values between the two groups, while the y-axis represents the statistical significance of that difference as –log10(p-value). CpG sites showing both high statistical significance and strong methylation differences are highlighted by color, especially those surpassing the Benjamini-Hochberg adjusted p-value threshold (shown in pink).  
  
Extract beta values (columns 1 to 8).
```{r}
beta <- final_ttest_corr[, 1:8]
```

Calculate mean beta-values for each group.

```{r}
beta_groupCTRL <- beta[, targets$Group == "CTRL"]
mean_beta_groupCTRL <- apply(beta_groupCTRL, 1, mean)

beta_groupDIS <- beta[, targets$Group == "DIS"]
mean_beta_groupDIS <- apply(beta_groupDIS, 1, mean)
```

Calculate delta (DIS - CTRL).

```{r}
delta_average <- mean_beta_groupDIS - mean_beta_groupCTRL
```

Identify probes significant after BH correction.

```{r}
BH_sig <- final_ttest_corr$corr_pval_BH < 0.01
```

Create a dataframe for the volcano plot.

```{r}
toVolcPlot <- data.frame(
  delta_average,
  minus_log10_p_val = -log10(final_ttest_corr$p_values_ttest),
  BH_sig
)
```

Display the volcano plot.

```{r}
plot(toVolcPlot[,1], toVolcPlot[,2], 
     pch=16, cex=0.4, col='grey16',
     xlab='Δμ_beta(DIS-CTRL)', ylab='-log10(p)',
     xlim=c(-0.8,0.8))
abline(h = -log10(0.01))

nominal_sig <- toVolcPlot[
  abs(toVolcPlot[,1]) > 0.1 & 
    toVolcPlot[,2] > -log10(0.01), ]
points(nominal_sig[,1], nominal_sig[,2], pch=16, cex=0.4, col="snow4")

BH_sig_points <- toVolcPlot[
  abs(toVolcPlot[,1]) > 0.1 & 
    toVolcPlot$BH_sig, ]
points(BH_sig_points[,1], BH_sig_points[,2], pch=19, cex=0.6, col="hotpink3")

legend("topright", 
       legend=c("not significant", "p < 0.01", "BH p < 0.01"),
       col=c("grey16", "snow4", "hotpink3"), pch = 19)
```

In this specific plot, no CpG sites reach the adjusted significance level, and thus no pink points are visible. This indicates that, although some sites show moderate differences, none pass the threshold for statistical significance after correction.  
  
**Manhattan Plot**  
In the Manhattan plot, the results of the same analysis are displayed across the genome. Each dot represents a CpG site, plotted according to its chromosomal position on the x-axis and its –log10(p-value) on the y-axis. This genome-wide overview enables the identification of differentially methylated regions (DMRs), highlighting clusters of epigenetic alterations that may play a role in disease processes.  
  
Add probe IDs as a column.

```{r}
final_ttest_corr_anno <- data.frame(rownames(final_ttest_corr), final_ttest_corr)
colnames(final_ttest_corr_anno)[1] <- "IlmnID"
```

Merge with Illumina manifest file (must contain IlmnID, CHR, MAPINFO).

```{r}
final_ttest_corr_anno <- merge(final_ttest_corr_anno, Illumina450Manifest_clean, by = "IlmnID")
```

Prepare the dataframe for `qqman` library.

```{r}
input_Manhattan <- data.frame(
  ID = final_ttest_corr_anno$IlmnID,
  CHR = final_ttest_corr_anno$CHR,
  MAPINFO = final_ttest_corr_anno$MAPINFO,
  PVAL = final_ttest_corr_anno$p_values_ttest
)
```

Convert chromosomes X and Y to numeric (23 and 24).

```{r}
levels(input_Manhattan$CHR)[levels(input_Manhattan$CHR) == "X"] <- "23"
levels(input_Manhattan$CHR)[levels(input_Manhattan$CHR) == "Y"] <- "24"
input_Manhattan$CHR <- as.numeric(as.character(input_Manhattan$CHR))
```

Load the `qqman` library and define a color palette  
qqman provides functions to generate Manhattan and QQ plots. It is useful for visualizing p-values in GWAS or methylation studies to detect significant loci.

```{r, warning=FALSE, message=FALSE, results='hide'}
install.packages("qqman")
library(qqman)
color_palette <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B", "#E377C2",
  "#7F7F7F", "#BCBD22", "#17BECF", "#AEC7E8", "#FFBB78", "#98DF8A", "#FF9896",
  "#C5B0D5", "#C49C94", "#F7B6D2", "#C7C7C7", "#DBDB8D", "#9EDAE5",
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728"
)
```

Display the Manhattan plot.

```{r}
manhattan(
  input_Manhattan,
  snp = "ID",
  chr = "CHR",
  bp = "MAPINFO",
  p = "PVAL",
  annotatePval = 0.00001,
  col = color_palette,
  suggestiveline = FALSE,
  genomewideline = -log10(0.00001)
)
```

CpG sites above the significance threshold (red line) are considered statistically significant after correction. Several top hits, such as cg07700747, cg08128400, and cg23388899, show strong association signals and may represent biologically relevant DMPs. The alternating colors per chromosome enhance readability and highlight the genomic distribution of significant probes.  
  
**12.	Produce an heatmap of the top 100 significant, differentially methylated probes.**  
  
The **heatmap** shows the methylation levels of selected CpG sites across all samples included in the study. Each row corresponds to a CpG site, and each column to a sample, with the color scale ranging from green (low methylation) to pink (high methylation). The samples are grouped by hierarchical clustering, revealing clear patterns of methylation that differentiate the CTRL and DIS groups. This clustering helps confirm the presence of consistent methylation signatures and supports the discriminative power of the selected sites. 
In this project, different clustering methods were applied: **single**, **complete**, and **average linkage**. These methods differ in how they define the distance between clusters.  
  
Load required library  
`gplots` is an R package that provides enhanced plotting functions for data visualization, including improved heatmaps, dendrograms, and tools for exploratory analysis of high-dimensional data.

```{r, warning=FALSE, message=FALSE, results='hide'}
if (!require("gplots")) install.packages("gplots")
library(gplots)
```

Prepare the input matrix for the heatmap.

```{r}
input_heatmap <- as.matrix(final_ttest[1:100, 1:8])
```

Assign colors to groups.

```{r}
group_color <- c()
for (i in 1:ncol(input_heatmap)) {
  name <- colnames(input_heatmap)[i]
  sample_id <- strsplit(name, "_")[[1]][1]
  matching_row <- which(targets$SampleID == sample_id)
  if (length(matching_row) > 0) {
    if (targets$Group[matching_row] == "CTRL") {
      group_color <- c(group_color, "#CBCBCB")
    } else {
      group_color <- c(group_color, "#393939")
    }
  }
}
```

Save Average linkage heatmap.

```{r}
heatmap.2(input_heatmap,
          col = terrain.colors(100),
          Rowv = TRUE,
          Colv = TRUE,
          hclustfun = function(x) hclust(x, method = 'average'),
          dendrogram = "both",
          key = TRUE,
          ColSideColors = group_color,
          density.info = "none",
          trace = "none",
          scale = "none",
          symm = FALSE,
          main = "Average linkage",
          key.xlab = 'beta-val',
          key.title = NA,
          keysize = 1,
          labRow = NA)
legend("topright",
       legend = levels(targets$Group),
       col = c('#CBCBCB', '#393939'),
       pch = 19,
       cex = 0.7)
```

Save Complete linkage heatmap.

```{r}
heatmap.2(input_heatmap,
          col = terrain.colors(100),
          Rowv = TRUE,
          Colv = TRUE,
          dendrogram = "both",
          key = TRUE,
          ColSideColors = group_color,
          density.info = "none",
          trace = "none",
          scale = "none",
          symm = FALSE,
          main = "Complete linkage",
          key.xlab = 'beta-val',
          key.title = NA,
          keysize = 1,
          labRow = NA)
legend("topright",
       legend = levels(targets$Group),
       col = c('#CBCBCB', '#393939'),
       pch = 19,
       cex = 0.7)
```

Save Single linkage heatmap.

```{r}
heatmap.2(input_heatmap,
          col = terrain.colors(100),
          Rowv = TRUE,
          Colv = TRUE,
          hclustfun = function(x) hclust(x, method = 'single'),
          dendrogram = "both",
          key = TRUE,
          ColSideColors = group_color,
          density.info = "none",
          trace = "none",
          scale = "none",
          symm = FALSE,
          main = "Single linkage",
          key.xlab = 'beta-val',
          key.title = NA,
          keysize = 1,
          labRow = NA)
legend("topright",
       legend = levels(targets$Group),
       col = c('#CBCBCB', '#393939'),
       pch = 19,
       cex = 0.7)
```

The three hierarchical clustering methods (Single, Complete, and Average linkage) show consistent global separation between CTRL and DIS samples, indicating robust methylation differences.   Complete and Average linkage provide more compact and biologically interpretable clusters, while Single linkage shows a chaining effect with elongated branches, which may reduce cluster resolution.  
Overall, Average linkage appears to balance sensitivity and interpretability, making it a suitable choice for downstream analysis.